Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-09 docs/manuscript/paper.tex
some manuscript updates
multiple noise realisations, shifted prior
rerun the canonical example, with a large dlogz
fixed h, multiple noise realisations, 1 core

\section{Signal tracking and parameter estimation} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Eq \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars. The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}. The synthetic residuals are generated by numerically solving Eq. \ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs. \ref{eq:sigmap}, \ref{eq:sigmap_f}. For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals. \textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo. Maybe just one is sufficient just as an example.}}
	\label{fig:qualitative_compare}
\end{figure}

@@

Initially we consider a single noise realisation of the synthetic data for the representative example. We apply the Kalman filter using the Earth terms only measurement equation, Eq. \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters. The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:corner_plot_1} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}. It is evident that for this representative example we are able to accurately estimate all the static parameters of the system. Regarding the estimates of $\boldsymbol{\theta}_{\rm gw}$, whilst all the parameters are generally recovered accurately, for some the posterior median is almost identically equal to the injected value  (e.g. $\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the  injected value (e.g. $\psi, \alpha$), whilst still being well within the distribution uncertainty. For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation. We will explore this further in Section \ref{sec:multiple_noise} and show that this is a bias that results from dropping the Earth terms, similar to the bias reported in \cite{Zhupulsarterms}. Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.

\begin{figure*}
	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/1237_broad_GW}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise. The vertical orange lines indicate the true injected values, c.f. Table \ref{tab:parameters_and_priors}. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
	\label{fig:example}%
\end{figure*}

@@

\subsubsection{Detection}
We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}. The Bayes factor, $\beta$, as defined in Eq. \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$. The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$. We can see that there is a general log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$. We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW. For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$. The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method. For these points the sampler converged sub-optimally for $\mathcal{M}_1$ and so the evidence $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)$ is lower than would be expected. This is purely a noise artefact; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these perturbations appear at different values of $h_0$, or else are not present at all. Increasing the number of live points $n_{\rm live}$ used by the nested sampler also dampens these perturbations and smooths the $\beta - h_0$ curve, since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$. Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails. That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds. This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.  
\begin{figure}
	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M_1}$ (GW present in data) and $M_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}. The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$. The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.}
	\label{fig:bayes}
\end{figure}

@@

The above discussion was just for a single realisaitn of the noise

@@

From Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$. Whilst both parameters are generally identifiable the algebraic relation between the two makes it difficult for our algorithm to estimate both simultaneously. For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$. This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$ but the algebraic relation of how these parameters get mapped into the measurement equation tends for our algorithm to trade off one parameter against the other. If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e. reduced the prior to a delta function). WE can see that the method can now infer $\iota$ which much greater accuracy.

@@

@@

@@

Points to discuss:

\begin{itemize}
	\item Non constant time sampling? Different pulsars sampled at different times...
\end{itemize}

>>>>

\section{Signal tracking and parameter estimation} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model outlined in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \ , \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N} \ ,
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
In Eq \ref{eq:params3} we have reparameterized the two GW polarisation strains, $h_{+}, h_{\times}$, in terms of the GW amplitude $h_0$ and the system inclination $\iota$:
\begin{align}
	h_+ &= h_0(1 + \cos^2 \iota) 	\label{eq:hphx} \ ,\\
	h_{\times} &= -2h_0\cos \iota 	\label{eq:hphx2} \ , 
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$, and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. We use this parametrisation in terms of $h_0$ and $\iota$ throughout the remainder of this work. For a PTA dataset containing $N$ pulsars we then have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters; for example estimates of pulsar distances are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}, but we have no prior information on the SMBHB GW source location.  \newline

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{images/example_residuals_plot}
	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars. The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}. The synthetic residuals are generated by numerically solving Eq. \ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs. \ref{eq:sigmap}, \ref{eq:sigmap_f}. For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals. \textcolor{red}{TK: potential sinusoidal feature? Need to discuss with Andres re use of baboo.}}
	\label{fig:qualitative_compare}
\end{figure}

@@

Initially we consider a single noise realisation of the synthetic data for the representative example. We apply the Kalman filter using the Earth terms only measurement equation, Eq. \ref{eq:measuremen_earth} in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters. The results are shown in Figure \ref{fig:corner_plot_1} for the 7 parameters of  $\boldsymbol{\theta}_{\rm gw}$ and in Figure \ref{fig:example_psr_params} for the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ in Figure \ref{fig:example}. It is evident that for this representative example we are able to accurately estimate all the static parameters of the system. Regarding the estimates of $\boldsymbol{\theta}_{\rm gw}$, whilst all the parameters are generally recovered accurately, for some the posterior median is almost identically equal to the injected value  (e.g. $\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the  injected value (e.g. $\psi, \alpha$), whilst still being well within the distribution uncertainty. For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation. We will explore this further in Section \ref{sec:multiple_noise} and show that this is a bias that results from dropping the Earth terms, similar to the bias reported in \cite{Zhupulsarterms}. Regarding $\boldsymbol{\theta}_{\rm psr, reduced}$, the frequency parameters $f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1)$ are generally well recovered for all pulsars, as are the noise amplitude parameters $\sigma^{(n)}$.

need some discussion here since htey are not perfecty recovered...

\begin{figure*}
	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/representative_example_v2_GW}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for a single realisation of the system noise. The vertical orange lines indicate the true injected values, c.f. Table \ref{tab:parameters_and_priors}. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

	\caption{As Fig \ref{fig:corner_plot_1} but for the $3N$ pulsar parameters $\boldsymbol{\theta}_{\rm psr, reduced}$. We show only a selection of the first 5 pulsars rather than the full PTA.}
	\label{fig:example_psr_params}%
\end{figure*}

@@

\subsubsection{Detection} \label{sec:detection}
We frame the problem of claiming a detection of a GW in the noisy data in terms of a Bayesian model selection procedure as described in Section \ref{sec:model_selection}. The Bayes factor, $\beta$, as defined in Eq. \ref{eq:bayes} is presented in Fig \ref{fig:bayes} for our representative system, where we now vary the magnitude of the GW strain, $h_0$. The noise processes in the synthetic data are identical realisations for each value of $h_0$; the only change in the data is the change in $h_0$. We can see that there is a general log-linear relationship between $\beta$ and $h_0$ where the GW source is easily detectable with decisive evidence ($\beta \sim 100$) for $h_0 \gtrapprox 10^{-14}$. We take $\beta = 10$ as a tolerance cut-off above which we can accept $\mathcal{M}_1$ and claim a detection of the GW. For this system, given this realisation of the noise, the minimal detectable strain is $\sim 3 \times 10^{-15}$. The small perturbations in the general log-linear $\beta - h_0$ trend arise due to noise in the nested sampling method. For these points the sampler converged sub-optimally for $\mathcal{M}_1$ and so the evidence $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1)$ is lower than would be expected. This is purely a noise artefact; if one recreates the $\beta - h_0$ curve by rerunning the nested sampler these perturbations appear at different values of $h_0$, or else are not present at all. Increasing the number of live points $n_{\rm live}$ used by the nested sampler also dampens these perturbations and smooths the $\beta - h_0$ curve, since as discussed in Section \ref{sec:nested_sampling} the uncertainties in $\mathcal{Z}$ scale as $\mathcal{O}(1/ \sqrt{n_{\rm live}})$. Below $\beta = 10$ this noise in the sampler starts to dominate and the hierarchical relationship between $\mathcal{M}_0$ and  $\mathcal{M}_1$ fails. That is, the requirement that $\mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_1) > \mathcal{Z}(\boldsymbol{Y} | \mathcal{M}_0)$ no longer holds. This is simply due to the fact that at these small strains the system is completely noise dominated and the two models $\mathcal{M}_1, \mathcal{M}_0$ cannot be separated.  
\begin{figure}
	\includegraphics[width=\columnwidth]{images/PaperBayesRatioPlotn1000_stacked}
	\caption{Bayes ratio $\beta$ between the competing models $\mathcal{M}_1$ (GW present in data) and $\mathcal{M}_0$ (GW not present in data) at different GW strain magnitudes $,h_0$, for the representative example summarised in Table \ref{tab:parameters_and_priors}. The horizontal grey dashed line labels the detection tolerance cut-off of $\beta = 10$. The minimal detectable strain, below which $\beta < 10$, is $\sim 3 \times 10^{-15}$.}
	\label{fig:bayes}
\end{figure}

@@

The above discussion was just for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.

To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.

For each realisation of the data we can then independently estimate the static parameters $\boldsymbol{\theta}$.

The estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 5 realisations of the noise are shown in Fig \ref{fig:corner_plot_2} which is analogous to Fig \ref{fig:corner_plot_1} but now shows multiple realisations of the noise.

We can see that the results agree well with the case of a single noise realisation. The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different realisations, agreeing with the injected values with high fidelity.

The variance in the estimates of $\iota$ and $h_0$ is much greater than for the other parameters.

This is due to the nested sampler effectively trading off accuracy in one parameter against the other; from Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.  For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$. This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$

Whilst both parameters are then generally identifiable, the specific algebraic relation between the two and how these parameters get mapped into the measurement equation makes it difficult for our algorithm to estimate both simultaneously, instead typically trading off one parameter against the other.

If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e. reduced the prior to a delta function). WE can see that the method can now infer $\iota$ which much greater accuracy.

Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.  Fig \ref{fig:example_psr_params}) show a consistency between the single and multiple noise realisations though for brevity we do not show the results here \textcolor{red}{TK: is this fair? Don't want to clog up paper with extra figures that don't offer any real insight.}

In the case where h is fixed...

whilst the estimates on  are much more varied.

In order to visualise the 1000 realisations, it is convenient to consider the distribution in the median of the posterior that is inferred for each static parameter, for each realisation of the noise. The resulting distributions are presented in Fig \ref{fig:median_distriubutins}

The bias that was evident for some of the parameters

  \textcolor{red}{with the estimate on iota}

are entirely consistent with the single

, which is analogous to Fig Y but with multiple  now shown.

\begin{figure*}
	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
	\caption{}
	\label{fig:corner_plot_2}
\end{figure*}

\begin{figure*}
	\setkeys{Gin}{width=\linewidth}   
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_omega_gw}
		\caption{$\Omega$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_phi0_gw}
		\caption{$\Phi_0$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_psi_gw}
		\caption{$\psi$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_iota_gw}
		\caption{$\iota$}
	\end{subfigure}

	\medskip
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_delta_gw}
		\caption{$\delta$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_alpha_gw}
		\caption{$\alpha$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_h}
		\caption{$h$}
	\end{subfigure}

	\caption{some caption \textcolor{red}{TK: these plots are ugly. Is there a better way to show this?}}\label{fig:median_distriubutins}
\end{figure*}

to generate 1000 different datasets

the above example was not simply a particualry

@@

@@

this could also go in "Multiple noise realisations"

@@

single pulsar performance.

@@

more stringent priors on f and fdot 
Points to discuss:

\begin{itemize}
	\item Non constant time sampling? Different pulsars sampled at different times...
	\item pulsar terms obv
\end{itemize}
